Types for R, Empirically Designed and Evaluated

### 2020-05-06 paper/main.tex
initial discussion of primitive types

\subsubsection{Types of Data}

@@

\section{Type System Design}

\AT{Lay out the type system.
Ideally, for every interaction we propose, we should be able to point to an analysis that supports our claim.
Maybe we don't always have to do this, especially for some of the simple, obvious types, but any departures from basic R functionality should be supported by the data.}

\section{Evaluation Results}

>>>>

\subsubsection{Types of Data}
\label{subsubsec:backgroundtypes}

@@

%
%
%
%
%
%
\section{Type System Design}
\label{sec:typesystemdesign}

Our corpora analysis reported \AT{number} call traces for over \AT{number} functions.
We distilled these call traces into function types which captured the dynamic behaviour of the functions.
We emphasize that this was an iterative process: initial analysis suggested a certain set of types, which we tried and implemented, and further analysis suggested improvements, and so on.
In this section, we will present the final type system that we designed, which we believe captures the essence of R.
Throughout, we will touch on design decisions and weigh our choices against alternatives, speaking to the strengths and weaknesses of our chosen design.

%
%
%
%
\subsection{Basic Types}
\label{subsec:basictypes}

As discussed in Section~\ref{subsubsec:backgroundtypes}, the workhorse types in R are the vectorized primitives: integers, doubles, complex numbers, logicals (booleans), characters, and raws (bytes).
These are {\it vectorized} in that they are always considered to be vectors by the R runtime: for instance, even ``scalar'' numbers are considered to be unit-length vectors to R.
These primitive vectors are homogeneous, in that a vector of doubles contains only doubles.

Related to these primitive types is one of R's notions of ``nullness'': for each of the primitives, R distinguishes a special ``NA'' value which represents missing data.
There is an NA for each of the primitive types (i.e. there is a double NA, integer NA, etc.).
Thus, even homogeneous vectors can represent missing data with their appropriate NA value.

As with everything in R, there is nuance in even the simplest cases.
We will now discuss our design decisions relating to the vectorized primitives.

%
%
\subsubsection{Possibly-NA Primitives}

A common primitive usage pattern is related to the presence of NAs: \AT{often}, programmers explicitly check for NAs in primitive vectors, possibly sanitizing them if they are present.
For instance, consider the code in \AT{do an NA example}.
NAs complicate practically every computation using primitive vectors, a fact that many programmers appear to be acutely aware of.

%
%
\subsubsection{Scalar Primitives}

\AT{Maybe we combine the next 3 subsections into one on dimensionality? There are overlapping arguments.}
Initial analyses of our data revealed that programmers \AT{often} use scalars (as explicitly as they can, anyway).
For example, consider the code in \AT{do a scalar example}.

We found that scalars occurred more often than vectors, and chose the scalar type syntax as the default.
The syntax for primitive scalars and vectors is as follows:

\AT{syntax...}

%
%
\subsubsection{Forgoing Matrices}

Another layer of complexity on vectorized primitives is the presence of ``matrices'', which are \AT{more detail}.
To R, a vector is a single-dimensional matrix, and R coerces vectors to matrices where appropriate.
That said, if all vectors are matrices, then any computation that is valid on a matrix should also be valid on a vector.
Restrictions on matrix-types arguments amount to restrictions on the dimensions of the argument, which suggests a dependent type system.
We will touch on this shortly.
\AT{I can try to come up with some numbers, here. I'm open to a discussion on how best to word this.}

%
%
\subsubsection{Forgoing Dimensionality}

\AT{Will revisit to tighten discussion, need to think about best way to phrase this.
Perhaps "dependent type system bad" is enough?}
Including data dimensions in types would make our type system a dependent type system, which we believe is too complex to retrofit onto a dynamic data science language such as R.
The simplest way to include dimensions would be to explicitly specify the dimensions of an argument (e.g. specifying that a function should always be called with a double vector of length 10), which is not particularly useful, and a pattern which \AT{rarely appeared in our analysis}.

%
%
%
%
\subsection{Lists}

%
%
%
%
%
%
\section{Evaluation Results}
